Class differences in social networks: Evidence from a referral experiment

### 2025-06-11 writing/manuscript.tex
draft 1 progress - results suite

\begin{figure}[H]
  \centering
  \caption{Network characteristics and courses taken together}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/share_same_program_by_tie.png}
      \caption{Same-program share}  
      \label{share_same_program_by_tie}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/conns_by_tie.png}
      \caption{Network size}
      \label{conns_by_tie}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} The left panel illustrates the share of connections within the same program as a function of the number of courses taken together. The right panel shows the average network size as a function of the number of courses taken together. Taking more than 5 courses together with a network member means on average 90\% chance to be in the same program. Similarly, past 5 courses together, the average network size dwindles by 80\%, from more than 220 individuals to below 50. 
  }
\end{figure}

\begin{figure}[H]
  \centering
  \caption{University programs by yearly fee and SES distribution}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/fees_with_boxplot.png}
      \caption{Programs sorted by fee}  
      \label{fees_with_boxplot}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/ses_distribution_by_fees.png}
      \caption{SES distribution by program fee}
      \label{ses_distribution_by_fees}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} The left panel shows the distribution of programs at the university by their average yearly fee. The right panel illustrates the distribution of each SES group across programs sorted by fee. As of 2025 net average monthly wage is around \$350 and the minimum legal wage is at \$200. The average yearly fee of programs stands at \$3000, and medicine is an extreme outlier at \$6000. Distributions of SES groups across programs show the majority of low-SES select into programs with below average cost, while high-SES select into programs with above average cost. Medicine accounts for a third of all high-SES students at this university.}
\end{figure}

@@

\section{Sample, Incentives, and Procedure} \label{s4}
We invited all 4,417 UNAB students who had at the time of recruitment completed their first year at the university to participate in our experiment. A total of 837 students took part in the data collection with a 19\% response rate. Our final sample consists of 734 individuals who referred peers with whom they have taken at least one class together, resulting in an 88\% success rate for the sample. We randomly allocated half of the participants into either \textbf{Baseline} or \textbf{Bonus} treatments. Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants took 3.8 courses together with their network, and the average network consisted of 175 peers.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active UNAB students who have SABER 11 scores in the administrative records. To manage budget constraints while maintaining sufficient incentives, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion, plus potential earnings from one randomly selected belief question (up to \$5) and one randomly selected recommendation question (up to \$60), for maximum total earnings of \$82. The random selection of payment part ensured that participants had incentives to exert effort across all tasks rather than focusing on a single part. Payment processing occured through online banking platform Nequi within 15 business days of participation.

\textcolor{cyan}{Data collection occurred during the last two weeks of April 2024. Our local partner at UNAB coordinated scheduled classroom visits and recruited research assistants to administer the experiment. Students present in class on the scheduled visit dates participated. Each classroom visit constituted a separate session. There were in total 35 sessions. Participants accessed the Qualtrics-based experiment using their smartphones during these visits. The median time to complete the survey was 20 minutes, with a compensation of \$26 for 117 lottery winners.}

@@

What are diversity related consequences of SES-driven differences across networks? In terms of network compositions, SES groups may connect at different rates with other SES groups than at random (Figure \ref{availability_lses}). Our results suggest at the network-level, SES groups form connections that mirror the overall university composition, with no significant deviations from expected proportions based on random sorting (all proportion tests have $p > 0.1$ across SES group comparisons). Each SES group connects with low, middle, and high SES peers at rates statistically indistinguishable from what would be expected given the demographic composition of the university.

@@

\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by time spent at UNAB}\label{share_same_ses_by_tie}
  \includegraphics[width=0.6\linewidth]{../figures/share_same_ses_by_tie.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure displays the average share of same-SES connections for low-, middle-, and high-SES as a function of the average number of courses taken together with network members. Low- and high-SES networks both become more homogenous as the average number of courses taken together with their connections increase.
  \end{tablenotes}
\end{figure}

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Distribution of referrals by area}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccc}
  \toprule
  Area & Only one referral & Both areas & Total \\
  \midrule
  Verbal & 65 & 608 & 673 \\
  Math & 61 & 608 & 669 \\
  \midrule
  Total & 126 & 1,216 & 1,342 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} The table shows how many referrers made referrals in only one area versus both areas. ``Only one referral'' indicates individuals who made referrals exclusively in that area. ``Both areas'' shows individuals who made referrals in both verbal and math areas. The majority of referrers (608) made referrals in both areas.
  \end{tablenotes}
  \label{tab:referral_distribution}
  \end{threeparttable}
\end{table}

@@

>>>>

@@

\section{Sample, Incentives, and Procedure} \label{s4}
We invited all 4,417 UNAB undergraduate students who had at the time of recruitment completed their first year at the university to participate in our experiment. A total of 837 students took part in the data collection with a 19\% response rate. Our final sample consists of 734 individuals who referred peers with whom they have taken at least one class together, resulting in an 88\% success rate for the sample. We randomly allocated half of the participants into either \textbf{Baseline} or \textbf{Bonus} treatments. Table \ref{tab:balance} presents key demographic characteristics and academic performance indicators across treatments (see Appendix Table \ref{tab:selection} for selection). The sample is well-balanced between the \textbf{Baseline} and \textbf{Bonus} conditions and we observe no statistically significant differences in any of the reported variables (all \textit{p} values $> 0.1$). Our sample is characterized by a majority of middle-SES students with about one-tenth of the sample being high-SES students. The test scores and GPA distributions are balanced. On average, participants took 3.8 courses together with their network, and the average network consisted of 175 peers.

@@

The experiment was conducted online through Qualtrics, with participants recruited from active UNAB students. To manage budget constraints while maintaining sufficient incentives, we randomly selected one in ten participants for payment. Selected participants received a fixed payment of \$17 for completion, plus potential earnings from one randomly selected belief question (up to \$5) and one randomly selected recommendation question (up to \$60), for maximum total earnings of \$82. The average time to complete the survey was 30 minutes, with an average compensation of \$80 for one in ten participants randomly selected for payment. Payment processing occured through online banking platform Nequi within 15 business days of participation.

@@

\subsection{SES diversity in networks}
What are diversity related consequences of SES-driven differences across networks? In terms of network compositions, SES groups may connect at different rates with other SES groups than at random (Figure \ref{availability_lses}). Our results suggest at the network-level, SES groups form connections that mirror the overall university composition, with no significant deviations from expected proportions based on random sorting (all proportion tests have $p > 0.1$ across SES group comparisons). Each SES group connects with low, middle, and high SES peers at rates statistically indistinguishable from what would be expected given the demographic composition of the university.

@@

At the same time, low-SES students connect with other low-SES students at significantly higher rates than high-SES students connect with low-SES students ($z = 2.05, p = .041$), while high-SES students show a marginally significant tendency to connect more with other high-SES students than low-SES students do ($z = -1.66, p = .098$). All other cross-group comparisons show no significant differences (all $p > .30$). Taken together, while our results show some SES segregation at group level for the extremes, these are balanced out, and do not point to problematic tendencies at the network-level.

So far we have looked at the entire network without considering the intensity of connections between students. In our network data set, this variable amounts to the number of classes taken together with peers. As we will see in the next section, referrals go to peers with whom participants have taken a lot of courses with \textcolor{red}{on average XX}, implying the intensity of the connection matters. We begin by dissecting what the intensity means in our context. As students take more courses together, the proportion of peers from the same academic program quickly goes beyond 95\% (see Figure \ref{share_same_program_by_tie}). Similarly, the average network size drops very quickly from above 210 to around 25 (see Figure \ref{conns_by_tie}). Both results indicate that actual referral considerations orginate from a much smaller pool of individuals from the same academic program.

\begin{figure}[H]
  \centering
  \caption{Network characteristics and courses taken together}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/share_same_program_by_tie.png}
      \caption{Same-program share}  
      \label{share_same_program_by_tie}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \centering
      \includegraphics[width=\linewidth]{../figures/conns_by_tie.png}
      \caption{Network size}
      \label{conns_by_tie}
  \end{subfigure}
  \caption*{\footnotesize\textit{Note:} The left panel illustrates the share of connections within the same program as a function of the number of courses taken together. The right panel shows the average network size as a function of the number of courses taken together. Taking more than 5 courses together with a network member means on average 90\% chance to be in the same program. Similarly, past 5 courses together, the average network size dwindles by 80\%, from more than 210 individuals to below 50. 
  }
\end{figure}

\subsection{SES diversity in networks}
What are the diversity implications of increasing the intensity of connections between students? As students take more courses together with peers, the share of same-SES peers in the networks of low- and high-SES increases while the share of middle-SES declines (see Figure \ref{share_same_ses_by_tie}). Both increases are substantial, amounting to 50\% for high-, and 30\% for low-SES. Combining these with the earlier result that beyond 5 courses taken together network members are almost entirely within the same program, these suggest program selection may have strong consequences for SES diversity in our setting.

\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by courses taken}\label{share_same_ses_by_tie}
  \includegraphics[width=0.6\linewidth]{../figures/share_same_ses_by_tie.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure illustrates the shares of same-SES connections for low-, middle-, and high-SES as a function of the average number of courses taken together with network members. Low- and high-SES networks both become more homogenous as the average number of courses taken together with their connections increase.
  \end{tablenotes}
\end{figure}

\subsection{Program selection and SES diversity}
Academic programs at this university are priced based on how much they cost, and typically less than 5\% of students receive any kind of scholarship \citep{diaz_peer_2025}. Based on these, we first calculate how much every program at the university is expected to cost students per year (see Figure \ref{fees_with_boxplot}). Considering that net minimum montly wage stands at \$200 and the average Colombian salary around \$350, the cost difference between programs are large enough to make an impact of program selection. Is it the case that SES groups select into programs with financial considerations?

\begin{figure}[H]
  \centering
  \caption{Programs sorted by fee}\label{fees_with_boxplot}
  \includegraphics[width=0.6\linewidth]{../figures/fees_with_boxplot.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure illustrates the distribution of programs at the university by their average yearly fee. The average yearly fee stands at \$3000, and medicine is an outlier at \$6000.
  \end{tablenotes}
\end{figure}

We look at how SES groups are distributed across programs to see evidence of SES-based selection (see Figure \ref{ses_distribution_by_fees}). Indeed, low-SES select into more affordable programs, followed by middle-SES. High-SES sort almost exclusively in above average costing programs, with a third of students selecting into medicine and creating a very skewed distribution. With this finding, program selection could be the reason why low- and high-SES networks tend to segregate as the number of courses taken increases. The next section characterizes the referrals, and we will return to the diversity implications of program selection once we understand how referrals were made.

\begin{figure}[H]
  \centering
  \caption{Programs sorted by fee}\label{ses_distribution_by_fees}
  \includegraphics[width=0.6\linewidth]{../figures/ses_distribution_by_fees.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure illustrates the distribution of each SES group across programs sorted by fee. the majority of low-SES select into programs with below average cost, while high-SES select into programs with above average cost. Medicine accounts for a third of all high-SES students at this university.
  \end{tablenotes}
\end{figure}

\subsection{Characterizing referrals}

We observe 1342 referrals in our final data set. More than 90\% of these consist from participants referring for both areas of the national entry exam, as can be seen in Table \ref{tab:referral_distribution}.

\begin{table}[H]
  \centering
  \begin{threeparttable}
  \caption{Distribution of referrals by area}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccc}
  \toprule
  Area & Only one area & Both areas & Total \\
  \midrule
  Verbal & 65 & 608 & 673 \\
  Math & 61 & 608 & 669 \\
  \midrule
  Total & 126 & 1,216 & 1,342 \\
  \bottomrule
  \end{tabular*}
  \begin{tablenotes}[flushleft]
  \footnotesize
  \item[] \textit{Note:} The table shows how many referrers made referrals in only one area versus both areas. ``Only one area'' indicates individuals who made referrals exclusively for one area of the exam. ``Both areas'' shows individuals who made referrals in both verbal and math areas. The majority of referrers (608) made referrals in both areas.
  \end{tablenotes}
  \label{tab:referral_distribution}
  \end{threeparttable}
\end{table}

What are the characteristics of the individuals who receive referrals, and how do they compare to others in the enrollment network?

@@

\subsection{Effect of the Bonus treatment}

\subsection{SES diversity in ex post referral choice sets}
At the median number of courses taken together for referrals, the effects of network segregation on SES diversity becomes clear (see Figure \ref{t15availability}) 
\begin{figure}[H]
  \centering
  \caption{Network size and courses taken together by courses taken}\label{t15availability}
  \includegraphics[width=0.6\linewidth]{../figures/t15availability.png}
  \begin{tablenotes}
    \footnotesize
    \item[] \textit{Note:} This figure illustrates the shares of same-SES connections for low-, middle-, and high-SES as a function of the average number of courses taken together with network members. Low- and high-SES networks both become more homogenous as the average number of courses taken together with their connections increase.
  \end{tablenotes}
\end{figure}
